Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-26 Report/Sections/intro.tex
Updates from Overleaf

{\em{This still needs some editing}}

\vskip 10pt\noindent

Recent advances in technology allow us to gather enormous amounts of data collected almost continuously in time.
The collection and analysis of movement data from devices such as accelerometers and GPS trackers have brought new understanding to {\bf{more references}}  physical activity levels in children \citep{Morris:2007} ...
In this article, we focus on the study of animal movement;  tags attached to animals record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours and providing biologists with new tools for understanding animal behaviour.

@@

Often, movement data can be viewed as a set of many curves.   In Morris, each child in the data set is associated with two curves from two different days.   In our data, we will consider curves from a sequence of dives of one killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada. 
These types of data are amenable to techniques of Functional Data Analysis (FDA).
FDA was originally developed for curves that can be considered independent replicates  (Ramsay ref, a theoretical ref).  
However, the structure of the data is often more complicated.
For instance, in Morris, clearly, the two curves from the same child are dependent.
And in our dive data, the sequence of days exhibits temporal dependence.

More complicated structures in FDA have been considered in the form of hierarchical ...  typically incorporating random effects.
Time series of curves  ...  RELATED would be appropriate here.
Here, we consider Markov modelling of a curve sequence.  Common in animal movement because it gives biological insight.

Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure via Markov processes.  
While our methodology is generally applicable, we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions, as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

The killer whale dives form a sequence of dependent curves. However, in FDA, curves are typically deemed independent.  
A noted exception involves multilevel modelling [references] which uses random effects and is appropriate when the set of curves can be partitioned into independent subsets, typically with each subset arising from one individual.  
[NEED REFERENCES]  
However, this approach is not appropriate here, so we consider a temporal structure for dependence, with the characteristics of each curve determined by a latent state that evolves according to a Markov model.  
This HMM allows us to interpret each curve as arising from a type of behaviour.

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.  Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

{\bf{ Make sure to reference Langrock}}

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017,Mcclintock:2020}.
One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. 
%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
%while maintaining the structure needed to run the forward algorithm. 
%It also only adds one additional parameter per possible hidden state.

Animal movement modelers have recognized that several simultaneous behavioral processes may occur at different time scales. 
One approach to the analysis of this type of data is to use a hierarchical hidden Markov model, or HHMM \citep{Barajas:2017,Adam:2019}. 
HHMMs model the entire time series in question as a nested structure of hidden Markov models where each HMM corresponds to one behavioral process.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. 
%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
One way to model periodic behavior in particular is to use signal processing techniques such as the Fourier transform on the raw data. 
The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. 
Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.

@@

>>>>

Recent advances in technology for tracking movement provide researchers with data collected almost continuously in time.
The collection and analysis of data from devices such as accelerometers have brought new understanding to areas ranging from  {\bf{more references}}  to physical activity levels in children \citep{Morris:2007}.
In this article, we focus on the study of animal movement;  tags attached to animals record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours and providing biologists with new information about  animal behaviour.

@@

Movement data are sometimes modelled as a set of curves that can  be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).  In \cite{Morris:2007}, a child's activity level throughout the day can be viewed as a curve, specifically, as activity as a function of time.   
In the data  we consider, a single dive of a whale can be viewed as a curve:  acceleration as a function of time.
FDA was originally developed for curves that can be considered independent replicates.
However, the structure of the data is often more complex.

Our goal is to develop movement analysis methods that provide subject-area insight  into the complex structure of curve data.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure and dependence from dive to dive via a hidden Markov model (HMM).
We model within dive variation by combining an HMM and a time series.
Using HMMs at both the dive level and the fine scale allows us to interpret the data as arising from different types of behaviour.
While our methodology is generally applicable, we  showcase our methods in a  study of the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive,  behaviour that seems to indicate some fine-scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

Particular types of between curve dependence structures have received attention in the FDA literature, 
notablly multilevel models that incorporate dependence via random effects.
For instance, in the aforementioned study of children's activity levels in  \cite{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
Perhaps the first work with this multilevel structure is \cite{Bromback:1998} in a study of ....   More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012}} and Crainiceanu [REF]. 
For our data, the multilevel models are not appropriate, since they do not account for the temporal dependence.

FDA researchers have modelled temporal dependence in a sequence of curves  via functional time series, extending the ideas of classic time series to model how one curve evolves into the next curve.  See, for instance, \cite{Kokoszka:2018}. 
While  functional time series methods can be applied to our dive data, they would not account for the different dive types.
Our HMM models account for both temporal dive-to-dive dependence and dive types.

%A noted exception involves multilevel modelling [references] which uses random effects and is appropriate when the set of curves can be partitioned into independent subsets, typically with each subset arising from one individual.

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.   Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017,Mcclintock:2020}, where
one of the most prevalent techniques in the literature of late is the HMM \citep{Patterson:2017}. 
However, HMMs have seen little use in FDA.  
One exception is the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioral processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here. 
Part of the purpose of our article is to introduce the FDA community to these very useful approaches to modelling complex curves.

%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
%while maintaining the structure needed to run the forward algorithm. 
%It also only adds one additional parameter per possible hidden state.

We also borrow techniques from animal movement to model fine-scale behaviour, beyond simply using an HMM.  
Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
Several solutions have been proposed, such as  the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.

An additional complication is that, at the shortest time scales,  observations often exhibit complicated  structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. 
%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
Our dive data exhibit  fine-scale periodic behaviour, so we transform the data via
 the signal processing technique of the Fourier transform. 
\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM  in the context of describing daily behavioral cycles of marine mammals.
In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. 
Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.

@@

In the following section, we describe HMMs, CarHMMs and HHMMs.  We show how these models can be combined in order to analyze complex data. We also discuss transforming data.  In Section 3, we study various models for complex ...  Section 4 contains data analysis and Section 5 discussion.

### 2020-07-26 Report/Sections/results.tex
Updates from Overleaf

In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts. However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

\subsection{Data Collection and Preprocessing}

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

\subsection{Model Selection}

We used a hierarchical HHM, where the coarse-scale observations were a time series of dive durations in seconds, and the fine-scale observations were the within-dive acceleration data (Figure \ref{fig:CarHHMM-DFT}). The coarse-scale level was comprised of an HMM with hidden states corresponding to dive types. The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale acceleration data exhibits significant sinusoidal behavior, thus the fine-scale observations $Y^*$ were transformed using the DFT summary statistics of a two-second sliding window. Unlike the simulation study, where the time series of acceration was one-dimensional, the acceleration data here was collected in three dimensions. These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. The following two sets of observations $z^*$ were calculated:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar. We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.

There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (section S-2.1), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM. We assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with the following parameters:
%
$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
%
where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.

While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear. Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$. In particular, the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean and variance:
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the subdive state $X_{t,t^*}$.

Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three subdive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.

\subsection{Results}

The estimates for the emission distribution parameters suggest that this whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting (CITATION, you can ask Ian and Sarah for help here - look at the Tennessen paper with HMM). Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with foraging behaviors in killer whales (Citation Tennessen maybe HMM and the other Tennesen with foraging data).

The three subdive behaviors also appeared distinct, particularly with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT},  Figure  \ref{fig:fine_emis}). Subdive type 1 corresponds to gliding and less overall activity compared to the other subdive types. The mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types. The auto-correlation of $\mathbf{Z}^{*(1)}$ also is higher for subdive type 1 than every other subdive type.
Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude high than subdive type 1. Acceleration has about twice the variance in this subdive type compared to subdive type 2, and slightly lower auto-correlation. Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior marine mammals \citep{Simon:2012}.
Finally, subdive type 3 corresponds to vigorous swimming activity, especially as the killer whale submerges or surfaces. The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher in subdive type 3 compared to every other state. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this subdive type, implying more variation in acceleration.

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix},$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix},$$
%
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \qquad
\hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
%
The coarse-scale stationary distribution indicates that most dives are short type 1 dives. The coarse-scale transition probability matrix further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active subdive type 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}. Such strategy could explain why subdive type 1 makes up 9\% of shallow dives for this killer whale, but it makes up 40\% of deep dives. Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 6 selected dives. The supplementary material also shows the estimated \textit{probability} of each dive time and subdive hidden state (section S-2.2).

@@

>>>>

%In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.

%\section{Results}

We report the results from fitting the CarHHMM-DFT in detail, validate the results, and then compare the results of this model with those from the other candidate models.

The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting (CITATION, you can ask Ian and Sarah for help here - look at the Tennessen paper with HMM).
Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with foraging behaviors in killer whales (Citation Tennessen maybe HMM and the other Tennesen with foraging data).

The three subdive behaviors also appeared distinct with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
For subdive type 1, the mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types. The auto-correlation of $\mathbf{Z}^{*(1)}$ also is higher for subdive type 1 than every other subdive type. This corresponds to less overall activity compared to the other subdive types. 
Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive type 1. Acceleration has about twice the variance in this subdive type compared to subdive type 1, and slightly lower auto-correlation. Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior in marine mammals \citep{Simon:2012}.
In subdive type 3, the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher compared to every other state. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this subdive type, implying more variation in acceleration. This corresponds to vigorous swimming activity, especially as the killer whale begins a dive or surfaces from a dive.

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix},$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix},$$
%
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
%
The coarse-scale stationary distribution indicates that most dives are short type 1 dives. The coarse-scale transition probability matrix further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active subdive type 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}. Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 6 selected dives. The supplementary material also shows the estimated probability of each dive and subdive type (section S-2.2).

@@

\subsection{Comparison with candidate models}

The CarHMM-DFT produced very similar to those of the CarHHMM-DFT on the fine scale. However, it failed to differentiate between short and long dives, and it therefore did not infer any information relating to the dive level Markov chain or the relationship between the dive level and the subdive level. For example, the CarHMM-DFT missed the insight that the whale was much more likely to be in subdive type 1 when engaged in longer dives than when engaged in shorter dives.

The HHMM-DFT decoded the dive and subdive types similarly to the CarHHMM-DFT, but was less likely to categorize the beginning and end of dives as subdive type 3. In addition, the HHMM-DFT produced estimates $\hat \sigma_1^*$ that were approximately 50-100 \% larger than those for the CarHHMM-DFT, and the estimated uncertainty of $\hat \mu^*_1$ was less than half of that for the CarHHMM-DFT for all dimensions and subdive states. These findings are consistent with the trend to overestimate variance and produce overconfident parameter estimates when auto-correlation is ignored. Further, the empirical auto-correlation of $\mathbf{Z}^{*(1)}$ within each decoded subdive type was at least 0.5 for all dimensions and subdive types. Finally, the pseudoresiduals of the HHMM-DFT had many more outliers than the pseudoresiduals of the CarHHMM-DFT, providing evidence that the HHMM-DFT was a worse fit to this data than the CarHHMM-DFT.

The CarHHMM did not have access to the Fourier sums of the acceleration data, so it failed to decode the subdive types into biologically interpretable results. In addition, the pseudoresiduals corresponding to the CarHHMM appears to be less heavy-tailed than a normal distribution.

See section S-??? for a more complete set of results for each of the candidate models.
